---
title: "Gerrymandering in Alabama"
author: "Gus Howard"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs/report") })
---


# Study metadata

Key Words: Compactness, Gerrymandering, Convex hull
Subject: Social and Behavioral Sciences: Geography: Geographic Information Sciences
Date: 02-18-2025
Spatial Coverage: Alabama
Spatial resolution: Census Block Groups
Spatial reference System: EPSG: 4269 NAD 1983 GCS
Temporal Coverage: 2020-2024
Temporal Resolution: Decennial Census


# Study design

This is an original study based on literature on gerrymandering metrics.

It is an exploratory study to evaluate the usefulness of a new gerrymandering metric based on
the convex hull of a congressional district and the representativeness inside the convex hull compared to the congressional district.


## Computational environment

I plan on using package the following packages: tidyverse - for general data manipulation, here - for file location, sf - for analyzing spatial data, tmap - to represent spatial data, tidycensus - for gathering census data, and kableExtra with R version 4.4.2.


```{r environment-setup, include = FALSE}
# record all the packages you are using here
# this includes any calls to library(), require(),
# and double colons such as here::i_am()
packages <- c("tidyverse", "here", "sf", "tmap", "tidycensus", "kableExtra", "lwgeom", "knitr", "htmltools", "markdown")

# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
# require(conflicted)

# load and install required packages
# https://groundhogr.com/

# if (!require(lwgeom)) {
#   install.packages("lwgeom")
#   require(lwgeom)
# }
# if (!require(tidyverse)) {
#   install.packages("tidyverse")
#   require(tidyverse)
# }
# if (!require(tidycensus)) {
#   install.packages("tidycensus")
#   require(tidycensus)
# }
# if (!require(tmap)) {
#   install.packages("tmap")
#   require(tmap)
# }
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}
# 
if (!require(here)) {
  install.packages("here")
  require(here)
}
# 
# if (!require(sf)) {
#   install.packages("sf")
#   require(sf)
# }
# 
# if (!require(markdown)) {
#   install.packages("markdown")
#   require(markdown)
# }
# 
# if (!require(kableExtra)) {
#   install.packages("kableExtra")
#   require(kableExtra)
# }
# 
# if (!require(knitr)) {
#   install.packages("knitr")
#   require(knitr)
# }

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
```

```{r}
groundhog.day <- "2025-03-08"

# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day)
# you may need to install a correct version of R
# you may need to respond OK in the console to permit groundhog to install packages
# you may need to restart R and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session

# record the R processing environment
# alternatively, use devtools::session_info() for better results
# writeLines(
#   capture.output(sessionInfo()),
#   here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
# )

# save package citations
knitr::write_bib(c(packages, "base"), file = here("software.bib"))

# set up default knitr parameters
# https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = FALSE, # Run code, show outputs (don't show code)
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

## Data Sources

### Precincts
- `Title`: Voting Precincts 2020
- `Abstract`: Alabama voting data for 2020 elections by precinct.
- `Spatial Coverage`: Alabama
- `Spatial Resolution`: Voting precincts
- `Spatial Reference System`: EPSG 4269 NAD 1983 geographic coordinate system
- `Temporal Coverage`: precincts used for tabulating the 2020 census
- `Temporal Resolution`: annual election

Precincts 2020
- VTDST20: Voting district ID
- GEOID20: Unique geographic ID
- G20PRETRU: Total vote for trump in 2020
- G20PREBID: Total votes for Biden in 2020

```{r, include = F}
precincts20 <- st_read(here("data","raw","public","districts.gpkg"), layer = "precincts20")


# includeMarkdown(here("data", "metadata", "precincts20.md"))
```



### Districts
- `Title`: Voting Districts 2020
- `Abstract`: Congressional Districts
- `Spatial Coverage`: Alabama
- `Spatial Resolution`: US Congressional Districts
- `Spatial Reference System`: EPSG 3857 WGS 1984 Web Mercator projection
- `Temporal Coverage`: Districts approved in 2023 for 2024
- `Temporal Resolution`: annual election

US Congressional districts
- DISTRICT: US Congressional District Number
- POPULATION: Total Population (2020 census)
- WHITE: total white population (2020 census)
- BLACK: total Black or African American population (2020 census)

```{r, include=FALSE}
districts23 <- st_read(here("data","raw","public","districts.gpkg"), layer = "districts23")

districts21 <- st_read(here("data","raw","public","districts.gpkg"), layer = "districts21")
```


### Block Groups
- `Title`: Block Groups 2020
- `Abstract`: Vector polygon geopackage layer of 2020 Census Block Groups and demographic data
- `Spatial Coverage`: Alabama
- `Spatial Resolution`: Census Block Groups
- `Spatial Reference System`: EPSG 4269 NAD 1983 geographic coordinate system
- `Temporal Coverage`: 2020 Census
- `Temporal Resolution`: 2020 Census
Block Groups
- GEOID: code to uniquely identify tracts
- P4_001N: Total Population, 18 years or older
- P4_006N: Total: Not Hispanic or Latino, Population of one race, Black or African American alone, 18 years or older
- P5_003N:Total institutionalized population in correctional facilities for adults, 18 years or older

```{r, include=FALSE}
census_metadata_file <- here("data", "metadata", "census2020pl_vars.csv")
if(file.exists(census_metadata_file)){
  census2020pl_vars <- read.csv(census_metadata_file)
} else {
  census2020pl_vars <- load_variables(2020, "pl")
  write.csv(census2020pl_vars, here("data", "metadata", "census2020pl_vars.csv"))
}
blockgroup_file <- here("data", "raw", "public", "block_groups.gpkg")

# if the data is already downloaded, just load it
# otherwise, query from the census and save
if(file.exists(blockgroup_file)){
  blockgroups <- st_read(blockgroup_file)
} else {
  blockgroups <- get_decennial(geography = "block group",
                               sumfile = "pl",
                               table = "P3",
                               year = 2020,
                               state = "Alabama",
                               output = "wide",
                               geometry = TRUE,
                               keep_geo_vars = TRUE)
  st_write(blockgroups, blockgroup_file)
}
```

transform block groups to WGS 1984

```{r}
blockgroups <- st_transform(blockgroups, 4269)
```

Map the districts

```{r}
tmap_mode(mode = "plot")
districts23map <- districts23 |> 
  tm_shape() +
  tm_polygons(fill_alpha = 0,
              col = "blue") +
  tm_labels(text = "DISTRICT",
          col="blue",
          bgcol = "white",
          bgcol_alpha = 0.5,
          on_surface = TRUE,
          just = c("center", "center")
          )

districts23map
```

## Prior observations  

At the time of this study pre-registration, the author had little prior knowledge of the geography of the study region with regards to the gerrymandering phenomena to be studied.
This study is related to one prior study by the author, in which the compactness of Alabama's voting districts was calculated, as well as voting statistics in relation to race.

## Bias and threats to validity

MAUP and Scale may affect these results.

Given the boundaries of this study are essential to the analysis, the Modifiable Areal Unit problem will certainly have relevant impacts. The constantly changing boundaries of voting precincts can impact both visualization of the data, as well as the analysis of it. The scale and shape of the units used has the potential to alter the data used in the study.

# Data Transformations

### Calculate Percent Black

```{r percent black}
black_vars <- census2020pl_vars |> 
  dplyr::filter(str_detect(name, "P3"),
                str_detect(label, "Black")) |> 
  select(-concept)

black_vars |> kable()
```

```{r, include = FALSE}
blockgroups_calc <- blockgroups |> 
  rowwise() |> 
  mutate(Black = sum(c_across(all_of(black_vars$name)))) |> 
  ungroup() |> 
  mutate(bgarea = st_area(geom),
         Total = P3_001N,
         PctBlack = Black / Total * 100,
         CheckPct = (Black + P3_003N) / Total * 100
         ) |> 
  select(GEOID, bgarea, Black, Total, PctBlack, CheckPct)
```

Save the results as `blockgroups_calc.gpkg`

```{r eval=FALSE}
st_write(blockgroups_calc, 
         here("data", "derived", "public", "blockgroups_calc.gpkg"),
         append=FALSE)
```


```{r, include=FALSE}
st_write(blockgroups_calc, 
         here("data", "derived", "public", "blockgroups_calc.gpkg"),
         append=FALSE)
```


Map the percentage of the population 18 or over that is Black or African American.

```{r}
blkgrp_black <- blockgroups_calc |> 
  tm_shape() + 
tm_shape(blockgroups_calc) + 
  tm_polygons(
    fill = "PctBlack",
    col_alpha = 0.2,
    lwd = 0.1,
    col = "grey90"
  )

blkgrp_black
```

## Analysis

First, this is the total population, but we should care more about the voting age population.
Second, we may want to categorize and calculate `BLACK` differently from the state of Alabama.


Spatial indices in R: https://r-spatial.org/r/2017/06/22/spatial-index.html

```{r warning=FALSE}
districts23_estimates <- st_intersection(blockgroups_calc, districts23) |> 
  mutate(
    awTot = Total * as.numeric(st_area(geom) / bgarea),
    awBlack = Black * as.numeric(st_area(geom) / bgarea)
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(bgTotal = sum(awTot),
            bgBlack = sum(awBlack))

districts23_join_bg <- districts23 |> 
  left_join(districts23_estimates, by = "DISTRICT") |> 
  mutate(pctBlackbg = round(bgBlack / bgTotal * 100, 1))
```

Report results.
We find very similar percentages of Black or African American people.

```{r}
districts23_join_bg |> st_drop_geometry() |> kable()
```


```{r warning=FALSE}
districts23_estimates <- st_intersection(blockgroups_calc, st_convex_hull(districts23)) |> 
  mutate(
    awTot = Total * as.numeric(st_area(geom) / bgarea),
    awBlack = Black * as.numeric(st_area(geom) / bgarea)
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(chTotal = sum(awTot),
            chBlack = sum(awBlack))
```

Join convex hull estimates to Districts with blockgroup estimates.

```{r}
districts23_join_ch <- districts23_join_bg |> 
  left_join(districts23_estimates, by = "DISTRICT") |> 
  mutate(pctBlackch = round(chBlack / chTotal * 100, 1),
         diffPct = pctBlackbg - pctBlackch,
         absdiffPct = abs(diffPct))
```



```{r}
districts23_results <- districts23_join_ch |> 
  mutate(
    darea = st_area(geom),
    dperim = st_length(st_cast(geom, "MULTILINESTRING")),
#    dperim2 = st_perimeter(geom),
    compact_shp = round( as.numeric((4 * pi * darea) / dperim^2), 2),
    compact_hull = round( as.numeric(darea / st_area(st_convex_hull(geom))), 2),
    compact_circ = round( as.numeric(darea / st_area(st_minimum_bounding_circle(geom))), 2)
  )
```




# Results

### Correlation matrix and small plots for gerrymandering indicators

```{r}
districts23_results_cor <- districts23_results |> 
  st_drop_geometry() |> 
  select(pctBlackbg,
         diffPct,
         absdiffPct,
         compact_shp,
         compact_hull,
         compact_circ)

districts23_results_cor |> cor() |> kable()
```

```{r}
districts23_results_cor |> pairs()
```



### Plot representational difference against compactness

Scatterplot with (absolute) difference in representation on x axis and compactness on y axis.
Plot the three different compactness scores simultaneously with different colors.
Symbolize the districts with different shapes.

```{r}
districts23_results_plot <- districts23_results |> 
  st_drop_geometry() |> 
  select(DISTRICT, pctBlackbg, absdiffPct, compact_shp, compact_hull, compact_circ) |> 
  pivot_longer(cols = starts_with("compact"))

districts23_results_plot |> ggplot() +
  aes(x = absdiffPct, y = value) +
  geom_smooth(method="lm", col = "grey30") +
  geom_label(aes(label = DISTRICT, fill = pctBlackbg)) +
  scale_fill_distiller(type = "div", palette = "PRGn") +

  facet_wrap(~name)
```



```{r}
tm_shape(districts23_results) +
  tm_polygons(fill = "pctBlackbg") +
  tm_text("DISTRICT")
```


# Discussion

Overall, there is a negative relationship between convex hull compactness and convex hull representational difference, as well as between minimum bounding circle compactness and convex hull difference. Notable exceptions include Districts 5 and 7. District 7 appears clearly gerrymandered—heavily packing African American voters—but the minimum bounding circle method fails to identify it as such. Conversely, District 5 is not meaningfully gerrymandered, yet the minimum bounding circle method flags it due to its elongated shape. Despite its length, District 5 remains relatively compact, which may explain the method's misclassification.

Compactness measures based on shape, convex hull, and minimum bounding circle generally correlate positively. However, District 5 is again an outlier in the relationship between shape-based and circle-based compactness.




This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, DOI: [10.17605/OSF.IO/W29MQ](https://doi.org/10.17605/OSF.IO/W29MQ)
